Pythran: Bringing OpenMP to a Python Subset

### 2013-02-11 doc/papers/europar2013/paper.tex
[Europar 2013] Few thoughts

To enable the user to write native functions, without having to write the glue
code that turns Python object into native structures back and forth, the SciPy
package provides the \texttt{weave} module that makes it possible to bundle C
code snippet into Python code and compile then load them at runtime.

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that can
communicate through IPC, using Python built-in object serialization. This
approach is only viable for computation intensive application. Furthermore, the
communication and synchronization overheads are much greater than a light-weight
processes approach.

Although the standard \texttt{threading} module makes it possible to start
several light-weight threads within the same interpreter, this approach is not
relevant for HPC, because of a specificity of CPython, the \emph{Global
Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
is active at a time in the interpreter. While it makes it possible to have
cooperative threads, say for a GUI, it does not take advantage of multiple
cores.  However, there are two notable exceptions: the GIL is released on I/O,
and the GIL does not prevent the use of threads inside native modules, where the
user has full control.

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. Its execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. So the GIL contention actually \emph{increases} the execution time.

@@

\subsubsection{Backward-Incompatible Approaches}

@@

The main issue of these approaches is that they imply to modify, in a more or
less complex way, the original code. They require to learn a new dialect,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code were to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into optimized ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

To be completely backward compatible, it must be possible to run the input code
in an environment that is not aware of the existence of the parallelizing
solution. This principle is not respected by parallel libraries, but code
annotations satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
annotations shine: the original code remains compatible with a compiler that is
not aware of OpenMP.

@@

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables that are referenced in a function
are considered when building such variable lists: there is no concept of
variable local to a block. Using the \texttt{default(none)} clause is possible
to ensure no variable gets forgotten when building such lists.

The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
characterize some data dependencies when performing a parallel reduction. The
list of supported operators depend on the input and backend languages: Python
has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distributes the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order. Integers used as loop
indices in Fortran and C satisfy this conditions, as well as C++ iterators with
the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
step of one until it is exhausted, in which case it raises an exception: it
behaves as a forward iterator. To be compatible with OpenMP, these iterators
must be turned into random access iterators.

@@

Generator expressions are generators whose content is built from an other
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random iterators only if their
input iterator is a random iterator itself.

@@

Pythran runtime is aware of these three kind of iterators: random access
iterators for standard containers, adaptor iterators for generator expressions
and forward iterators for generators.

@@

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used to detect when OpenMP is not available. Python does
not have a preprocessor, but it is possible to catch the import exception if the
\texttt{omp} module is not found.

@@

Pythran is a subset of Python designed for scientific computing. It is
implicitly statically typed and supports most Python constructs except those
that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
is currently only partially supported. The associated compiler turns Pythran
code, possibly annotated with OpenMP and a few function type annotations, into
C++ code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
relies on type inference to remove most dynamic behavior, combined with a
high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
between Python and C++ constructs. The purpose of this section is not to dig
into the internals of the compiler, but rather to focus on the impact of the
parallelism layer, especially on the runtime library.

@@

Pythran runtime is based on the STL. The standard library of C++11, unlike its
C++03 version, is thread-safe, so the data dependencies in the original code
remain explicit.

@@

Pythran handles the problem by refusing recursive types, which makes it possible
to solely rely on reference counting for memory management. It can be
implemented through the thread-safe shared pointer mechanism provided in C++11
standard library.

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still implies copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this overhead.

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, Hyantes,
we successively parallelized it with OpenMP, turned it into Python with a
Pythran compatible kernel and parallelized the Python version using the approach
described in this paper. We also measure the SLOC of the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It not
only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

The performance of the two approaches is shown in
Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
mathematical, image-processing or linear algebra kernels. All these kernels
have been written in Cython and Python ---compatible with Pythran--- and
annotated through the mechanism of each language to exhibit parallelism, then
compiled into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.
Their execution time when called from the Python interpreter is measured
through the \texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.
All results are normalized against Cython sequential execution time. They show
that while being of a higher level than Cython, Pythran achieves comparable
results.

@@

\begin{figure}[ht]
    \includegraphics[width=\textwidth]{cython}
    \caption{Comparison of Cython en Pythran generated code performance.}
    \label{fig:cython-pythran}
\end{figure}

>>>>

To enable the user to write native functions, without having to write the glue
code that turns Python object into native structures back and forth, the SciPy
package provides the \texttt{weave} module that makes it possible to bundle C
code snippet into Python code and compile then load them at runtime.
\marginpar{Strange to have a "partial" related work here, while weave is not even mention later (see further marginpar).}

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that can
communicate through IPC, using Python built-in object serialization. This
approach is only viable for computation intensive application, since the
communication and synchronization overheads are much greater than a light-weight
processes approach.

Although the standard \texttt{threading} module makes it possible to start
several light-weight threads within the same interpreter, this approach is not
applicable to HPC, because of a specificity of CPython, the \emph{Global
Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
IronPython or Jython, do not have a GIL.} 
\marginpar{The fact that others Python implementations do not have a GIL make all the argumentation dubious, especially the last paragraph.}
This lock ensures that only one thread
is active at a time in the interpreter. While it makes it possible to have
cooperative threads, say for a GUI, it does not take advantage of multiple
cores.  However, there are two notable exceptions: the GIL is released on I/O,
and the GIL does not prevent the use of threads inside native modules, where the
user has full control.

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. Its execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. So the GIL contention actually \emph{increases} the execution time.
\marginpar{MA: shortcut! What proves that the issue is the GIL? You may just have written bad code...}

@@

\subsubsection{Backward-Incompatible Approaches}
\marginpar{Why weave does not fit this section and is mention in the introduction.}

@@

The main issue of these approaches is that they imply to modify, in a more or
less complex way, the original code.\marginpar{just as pythran} They require to learn a new dialect,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code were to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.
\marginpar{JIT is never defined.}

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into optimized ones \marginpar{"optimized"? I would rather say "native"/"bytecode"}, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

To be completely backward compatible, it must be possible to run the input code
in an environment that is not aware of the existence of the parallelizing
solution\marginpar{First sentence is unclear to me.}. This principle is not respected by parallel libraries, but code
annotations satisfy it \marginpar{It is obviously not true for OpenMP. I don't see why it is not for parallel libraries!}. This is where OpenMP~\cite{openmp3.1} parallel
annotations shine: the original code remains compatible with a compiler that is
not aware of OpenMP.\marginpar{This is wrong.}

@@

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables that are referenced in a function
are considered when building such variable lists: there is no concept of
variable local to a block. Using the \texttt{default(none)} clause is possible
to ensure no variable gets forgotten when building such lists.
\marginpar{I'm not sure what's the point with this paragraph. Just drop it?}

The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
characterize some data dependencies when performing a parallel reduction. The
list of supported operators depend on the input and backend languages: Python
has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
\marginpar{Unclear to me what you're trying to explain. Python doesn't have \&\&, how is it relevant? C/C++ does not have  By the way OpenMP 4.0 has "user-defined" reduction operator.}

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distributes the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order.\marginpar{Is total order a real need?} Integers used as loop
indices in Fortran and C satisfy this conditions, as well as C++ iterators with
the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
step of one until it is exhausted, in which case it raises an exception: it
behaves as a forward iterator. To be compatible with OpenMP, these iterators
must be turned into random access iterators.

@@

Generator expressions are generators whose content is built from an other
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random iterators only if their
input iterator is a random iterator itself.\marginpar{Necessary condition, but is a sufficient one?}

@@

Pythran runtime is aware of these three kind of iterators: random access
iterators for standard containers, adaptor iterators for generator expressions
and forward iterators for generators.
\marginpar{This sentence is like a "cheveu sur la soupe", the whole section talks about parallelism issue, while this sentence is about language restriction in general?}

@@

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used to detect when OpenMP is not available. Python does
not have a preprocessor, but it is possible to catch the import exception if the
\texttt{omp} module is not found.
\marginpar{omp module? First time you're talking about that. How is involved? Is it standard? ...}

@@

Pythran is a subset of Python designed for scientific computing. It is
implicitly statically typed and supports most Python constructs except those
that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
(e.g.\ \texttt{eval}).\marginpar{class?} A few standard modules are supported in addition to the
core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
is currently only partially supported. The associated compiler turns Pythran
code, possibly annotated with OpenMP and a few function type annotations, into
C++ code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
relies on type inference to remove most dynamic behavior, combined with a
high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
between Python and C++ constructs. The purpose of this section is not to dig
into the internals of the compiler, but rather to focus on the impact of the
parallelism layer, especially on the runtime library.

@@

Pythran runtime is based on the STL. The standard library of C++11, unlike its
C++03 version, is thread-safe, so the data dependencies in the original code
remain explicit.
\marginpar{I've never seen "explicit" dependencies in my C++ code, what's that?}

@@

Pythran handles the problem by refusing recursive types, which makes it possible
to solely rely on reference counting for memory management. It can be
implemented through the thread-safe shared pointer mechanism provided by the C++11
standard library.

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still implies copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this overhead.
\marginpar{what is the point about reference or const refence? I mean: who cares? Nobody reads the generated code...
Moreover, what about copy elision made by the compiler? You're tied to a single compilation unit so the C++ compiler should be able to make it anyway.}

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, Hyantes,
we successively parallelized it with OpenMP, turned it into Python with a
Pythran compatible kernel and parallelized the Python version using the approach
described in this paper. We also measure the SLOC of the two versions. \marginpar{unfair, and arguably relevant.}
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It not
only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

The performance of the two approaches is shown in
Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
mathematical, image-processing or linear algebra kernels. All these kernels
have been written in Cython and Python ---compatible with Pythran--- and
annotated through the mechanism of each language to exhibit parallelism, then
compiled into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.
Their execution time when called from the Python interpreter is measured
through the \texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.
All results are normalized against Cython sequential execution time. They show
that while handling code at a higher level than Cython, Pythran achieves comparable
results.

@@

\begin{figure}[ht]
    \includegraphics[width=\textwidth]{cython}
    \caption{Comparison of Cython and Pythran generated code performance.}
    \label{fig:cython-pythran}
\end{figure}
